A general Nullstellensatz for generalized spaces

### 2020-01-21 paper-qcoh.tex
to do
typo

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, the theory of local rings or the theory of intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$). By (infinitary) \emph{first-order
formula} we will mean a formula which may contain, additionally to the
connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
and~${\forall}$.

@@

One of the most celebrated such sequents arises in the case that~$\TT$ is the
theory of local rings. In this case, the classifying topos~$\Set[\TT]$ is also
known as the \emph{big Zariski topos} of~$\Spec(\ZZ)$ from algebraic geometry,
the topos of sheaves over the site of schemes locally of finite presentation,
and the generic model is the functor~$\affl$ of points of the affine line, the
functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom_\mathrm{Sch}(X,\mathbb{A}^1) =
\O_X(X)$.

@@

\paragraph{A general Nullstellensatz} To explain the relevant background, the
somewhat vague question ``to which extent does the classifying
topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
useful guiding principle. This is easiest to visualize with a concrete example
for~$\TT$, such as the theory of rings.

@@

Given an internal signature~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
elementary topos with a natural numbers object), we can successively build the
object of contexts (the object of lists of sorts); the object of terms
(equipped with a morphism to the object of contexts); the object of atomic
propositions (again equipped with such a morphism); the object of geometric
formulas (again so); and the object of geometric sequents (again so). A internal
geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
of geometric sequents, interpreted as the object of axioms of~$\TT$.
Given such an internal theory~$\TT$, we can build the subobject of the provable
sequents.

@@

Disjunctions appearing in internal geometric formulas may be indexed by
arbitrary objects of the topos (that is, disjunctions may be taken over
arbitrary subobjects of disjuncts), just like disjunctions appearing in ordinary
external geometric formulas over an ordinary signature may be indexed by
arbitrary sets. If~$\Sigma$ is an internal signature in a Grothendieck topos~$\E$, the object of geometric
formulas over~$\Sigma$ has an important
subobject, the subobject of those formulas such that locally, any appearing
disjunction is indexed by a constant sheaf. Such internal geometric formulas
will be called \emph{geometric$^\star$ formulas}.

@@

\begin{scholium}\label{scholium:nullstellensatz-more-specific}
Let~$\TT$ be a geometric theory. Let~$\vec x$ be a context over the signature
of~$\TT$. Let~$(\sigma_i)_i$ be a set of geometric sequents, where each context
is of the form~$\vec x, \vec y_i$ for some additional list~$\vec y_i$.
Write~$\sigma_i = (\varphi_i \seq{\vec x, \vec y_i} \psi_i)$. Then, internally
to~$\Set[\TT]$, it holds that
\[ \forall \vec x\_
  \bigl(\bigwedge_i (\forall \vec y_i\_ \varphi_i \Rightarrow \psi_i)\bigr) \Longrightarrow
  \bigvee_\alpha \alpha,
\]
where the disjunction ranges over all those geometric formulas~$\alpha$ in the
context~$\vec x$ such that for each index~$i$, the theory~$\TT$ proves~$(\alpha
\wedge \varphi_i \seq{\vec x, \vec y_i} \psi_i)$.
\end{scholium}

@@

\begin{proof}[Proof of Theorem~\ref{thm:characterization}]
Any infinitary first-order formula can be simplified to an equivalent geometric
formula -- on the semantic side by repeatedly applying
Scho\-li\-um~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
reduced to the basic fact (Proposition~\ref{prop:basic-truth}) that, for
geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ if and only if~$\TT$ proves~$\varphi$.
\end{proof}

@@

Throughout this section, let~$\TT$ be a Horn theory. It is a basic result that
Horn theories are of presheaf type, such that the classifying topos~$\Set[\TT]$
can be taken as the topos of functors~$\Mod{\TT}_\mathrm{fp} \to \Set$,
where~$\Mod{\TT}_\mathrm{fp}$ is the full subcategory of the category
of~$\TT$-models (in~$\Set$) on the finitely presentable objects~\cite[Theorem~2.1.21]{caramello:tst}. The
(underlying object of the) generic model is the tautologous functor~$U_\TT : T
\mapsto T$. By~\emph{$U_\TT$-algebra} we mean, in analogy with the terminology in
the case that~$\TT$ is the theory of rings, a~$\TT$-model~$M$ equipped with
a~$\TT$-homomorphism~$U_\TT \to M$.

@@

\begin{proof}The desired model can be constructed as a term algebra. As a set,
it consists of the terms (in the empty context) of the signature~$\Sigma + X$
modulo the equivalence relation identifying two terms if and only if~$\TT + R$
proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
\brak{S} \Leftrightarrow (\TT + R \,\vdash S(t_1,\ldots,t_n))$.

@@

\begin{proof}Limits are computed as the limits of the underlying sets, colimits
are computed by using the construction of Lemma~\ref{lemma:free-models}. For
instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}

@@

Following the Kripke--Joyal translation of this statement, let a stage~$T \in
\Mod{\TT}_\mathrm{fp}$,~$T$-elements~$X_1,\ldots,X_n$ of the object of sorts of
the signature of~$\ul{\TT}/U_\TT$ (that is the constant sheaf on the set of
sorts of~$\TT$), and~$T$-elements~$R_1,\ldots,R_m$ of the object of atomic
propositions over the signature of~$\ul{\TT}/U_\TT$ be given. By passing to a
covering of~$T$, we may assume that the~$X_i$ are given by sorts of~$\TT$ and
that the~$R_j$ are given by atomic propositions over the signature
of~$\TT/U_\TT(T)$.

@@

Butz and Johnstone~\cite{butz-johnstone:first-order} list several nongeometric sequents holding in the
classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
no function symbols, relations symbols or axioms, including:
\begin{align*}
  \forall x,y\?U_\OO\_ \neg\neg(x = y) &&
  \forall x_1,\ldots,x_n\?U_\OO\_ \neg \forall y\?U_\OO\_ \bigvee_{i=1}^n (y = x_i)
\end{align*}
Intuitively, the first expresses that~$U_\OO$ is close to a subsingleton, while
the second expresses that~$U_\OO$ is close to being infinite. Using the
Nullstellensatz, these formulas can be verified as follows.

>>>>

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, the theory of local rings or the theory of intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
% XXX: The terminology is not due to Olivia, right? Find better expression
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$). By (infinitary) \emph{first-order
formula} we will mean a formula which may contain, additionally to the
connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
and~${\forall}$.

@@

One of the most celebrated such sequents arises in the case that~$\TT$ is the
theory of local rings. In this case, the classifying topos~$\Set[\TT]$ is also
known as the \emph{big Zariski topos} of~$\Spec(\ZZ)$ from algebraic geometry,
the topos of sheaves over the site of schemes locally of finite presentation,
and the generic model is the functor~$\affl$ of points of the affine line, the
functor which maps any scheme~$X$ (l.o.f.p.) to~$\Hom_\mathrm{Sch}(X,\mathbb{A}^1) =
\O_X(X)$.

@@

\paragraph{A general Nullstellensatz} To explain the relevant background, the
somewhat vague question ``to which extent does the classifying
topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is useful as
a guiding principle. This is easiest to visualize with a concrete example
for~$\TT$, such as the theory of rings.

@@

Given an internal signature~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
elementary topos with a natural numbers object), we can successively build the
object of contexts (the object of lists of sorts); the object of terms
(equipped with a morphism to the object of contexts); the object of atomic
propositions (again equipped with such a morphism); the object of geometric
formulas (again so); and the object of geometric sequents (again so). An internal
geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
of geometric sequents, interpreted as the object of axioms of~$\TT$.
Given such an internal theory~$\TT$, we can build the subobject of the provable
sequents.

@@

Disjunctions appearing in internal geometric formulas may be indexed by
arbitrary objects of the topos, just like disjunctions appearing in ordinary
external geometric formulas over an ordinary signature may be indexed by
arbitrary sets. If~$\Sigma$ is an internal signature in a Grothendieck topos~$\E$, the object of geometric
formulas over~$\Sigma$ has an important
subobject, the subobject of those formulas such that locally, any appearing
disjunction is indexed by a constant sheaf. Such internal geometric formulas
will be called \emph{geometric$^\star$ formulas}.

@@

\begin{scholium}\label{scholium:nullstellensatz-more-specific}
Let~$\TT$ be a geometric theory. Let~$\vec x$ be a context over the signature
of~$\TT$. Let~$(\sigma_i)_i$ be a set of geometric sequents, where each context
is of the form~$\vec x, \vec y_i$ for some additional list~$\vec y_i$.
Write~$\sigma_i = (\varphi_i \seq{\vec x, \vec y_i} \psi_i)$. Then, internally
to~$\Set[\TT]$, it holds that
\[ \forall \vec x\_
  \bigl(\bigwedge_i (\forall \vec y_i\_ \varphi_i \Rightarrow \psi_i)\bigr) \Longrightarrow
  \bigvee_\alpha \alpha,
\]
where the disjunction ranges over all those geometric formulas~$\alpha$ in the
context~$\vec x$ such that for every index~$i$, the theory~$\TT$ proves~$(\alpha
\wedge \varphi_i \seq{\vec x, \vec y_i} \psi_i)$.
\end{scholium}

@@

\begin{proof}[Proof of Theorem~\ref{thm:characterization}]
Any infinitary first-order formula can be simplified to an equivalent geometric
formula -- on the semantic side by repeatedly applying
Scho\-li\-um~\ref{scholium:nullstellensatz-more-specific}, on the syntactic side by
repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
reduced to the basic fact (Proposition~\ref{prop:basic-truth}) that, for
geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ if and only if~$\TT$ proves~$\varphi$.
\end{proof}

@@

Throughout this section, let~$\TT$ be a Horn theory. It is a basic result that
Horn theories are of presheaf type, such that the classifying topos~$\Set[\TT]$
can be taken as the topos of functors~$\Mod{\TT}_\mathrm{fp} \to \Set$,
where~$\Mod{\TT}_\mathrm{fp}$ is the full subcategory of the category
of~$\TT$-models (in~$\Set$) on the finitely presentable objects~\cite[Theorem~2.1.21]{caramello:tst}. The
% XXX: besser formulieren family of objects
(underlying object of the) generic model is the tautologous functor~$U_\TT : T
\mapsto T$. By~\emph{$U_\TT$-algebra} we mean, in analogy with the terminology in
the case that~$\TT$ is the theory of rings, a~$\TT$-model~$M$ equipped with
a~$\TT$-homomorphism~$U_\TT \to M$.

@@

\begin{proof}The desired model can be constructed as a term algebra. As a set,
it consists of the terms (in the empty context) of the signature~$\Sigma + X$
modulo the equivalence relation identifying two terms if and only if~$\TT + R$
proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
relation symbols~$H$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
\brak{H} \Leftrightarrow (\TT + R \,\vdash H(t_1,\ldots,t_n))$.

@@

\begin{proof}Limits are computed as the limits of the underlying sets, colimits
are computed by using the construction of Lemma~\ref{lemma:free-models}. For
instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
% XXX sagen, dass jedes Modell (beliebig) präsentiert wird

@@

Following the Kripke--Joyal translation of this statement, let a stage~$T \in
\Mod{\TT}_\mathrm{fp}$,~$T$-elements~$X_1,\ldots,X_n$ of the object of sorts of
the signature of~$\ul{\TT}/U_\TT$ (that is the constant sheaf on the set of
sorts of~$\TT$), and~$T$-elements~$R_1,\ldots,R_m$ of the object of atomic
propositions over the signature of~$\ul{\TT}/U_\TT$ be given. By passing to a
covering of~$T$, we may assume that the~$X_i$ are given by sorts of~$\TT$ and
that the~$R_j$ are given by atomic propositions over the signature
of~$\TT/U_\TT(T)$.
% XXX: No passing to covers is necessary, as we're in a *pre*sheaf topos!

@@

Butz and Johnstone~\cite{butz-johnstone:first-order} list several nongeometric sequents holding in the
classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
no function symbols, relation symbols or axioms, including:
\begin{align*}
  \forall x,y\?U_\OO\_ \neg\neg(x = y) &&
  \forall x_1,\ldots,x_n\?U_\OO\_ \neg \forall y\?U_\OO\_ \bigvee_{i=1}^n (y = x_i)
\end{align*}
Intuitively, the first expresses that~$U_\OO$ is close to a subsingleton, while
the second expresses that~$U_\OO$ is close to being infinite. Using the
Nullstellensatz, these formulas can be verified as follows.
